%========================
% paper.tex
%========================
\documentclass[11pt]{article}

%---- Page & fonts
\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[american]{babel}
\usepackage{microtype}
\usepackage{authblk}

%---- Math & symbols
\usepackage{amsmath, amssymb, mathtools}
\usepackage{bbm}          % for \mathbbm{1} indicator
\usepackage{siunitx}

%---- Graphics, tables, floats
\usepackage{graphicx}
\graphicspath{{fig/}{figures/}{build/}} % adjust as needed
\usepackage{booktabs, threeparttable}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{placeins}

%---- Links & clever refs
\usepackage[hidelinks]{hyperref}
\usepackage[capitalize,noabbrev]{cleveref}

%---- Citations & bib
\usepackage[
  backend=biber,       % requires biber
  style=authoryear,    % (Author Year)
  maxcitenames=2,
  maxbibnames=99,
  uniquename=false,
  dashed=false
]{biblatex}
\addbibresource{/Users/hectorbahamonde/Bibliografia_PoliSci/library.bib}  % your .bib file


%---- Custom macros (edit as you like)
\newcommand{\govdist}{\texttt{gov\_distance\_w\_01}}
\newcommand{\govclose}{\texttt{gov\_closeness\_w\_01}}
\newcommand{\techno}{\texttt{techno5}}
\newcommand{\demsat}{\texttt{Q8\_1}}      % democratic satisfaction
\newcommand{\trustpol}{\texttt{Q9\_4}}    % trust in Finnish politicians
\newcommand{\ind}[1]{\mathbbm{1}\!\left\{#1\right\}}

\title{Losers, Delegation, and the Ideological Valence of Expertise:\\ Evidence from Finland}


\author[1]{%
\textsc{H\'ector Bahamonde}%
\thanks{\href{mailto:hector.bahamonde@utu.fi}{hector.bahamonde@utu.fi}; %
\href{http://www.hectorbahamonde.com}{\texttt{www.HectorBahamonde.com}}.}%
}

\author[2]{%
\textsc{Inga Saikkonen}%
\thanks{\href{mailto:inga.saikkonen@abo.fi}{inga.saikkonen@abo.fi}; %
\href{https://research.abo.fi/en/persons/inga-saikkonen}{\texttt{research.abo.fi/en/persons/inga-saikkonen}}.\\
Authors are listed in alphabetical order and both have contributed equally to the paper.}%
}

\affil[1]{Senior Researcher, University of Turku, Finland}
\affil[2]{Senior University Lecturer, {\AA}bo Akademi University}

\date{\today}

\begin{document}
\maketitle

\begin{abstract}
Polarization research shows that voters often trade democratic procedures for partisan ends, weakening electoral checks on incumbents. Yet we know far less about how citizens evaluate \emph{delegation}—shifting authority from elected politicians to unelected experts. We bridge this gap by theorizing that delegation is not ideologically neutral: its appeal depends on the perceived ideological valence of “expert” institutions. We argue that electoral losers will oppose delegation when experts are seen as aligned with the incumbent coalition and may favor it when experts are perceived as neutral or aligned with the opposition. Empirically, we leverage a novel, original survey from Finland (2023–2025), a high-quality, multiparty parliamentary democracy—an analytically demanding hard case where expert governance might otherwise be broadly acceptable. We develop an individual-level, seat-weighted measure of distance to the governing coalition based on party-closeness evaluations (renormalized for item nonresponse) and study its association with support for delegating power to experts. Linear and ordered-logit models include demographic covariates, democratic satisfaction, trust in politicians, and region fixed effects; we visualize average predicted probabilities across the full distance range. Contrary to a simple “losers favor constraints” expectation, greater distance from the right-leaning cabinet is associated with \emph{lower} support for technocratic delegation. Substantively, this identifies when losers’ consent does—and does not—extend to technocracy; methodologically, we introduce a transparent distance-to-government metric that travels to coalition systems.
\end{abstract}

\newpage

\section{Introduction}

Democratic backsliding increasingly advances through legalistic tactics rather than overt rupture, and partisan trade–offs often blunt voters’ willingness to discipline incumbents \parencite{Svolik2020QJPS, Svolik2019JoD}. Classic work links party competition and polarization to regime vulnerability \parencite{Sartori1976, Dahl1971, Lipset1960}, while recent studies document how incumbents tilt the playing field without abolishing elections \parencite{LevitskyWay2010, Bermeo2016, WaldnerLust2018}. This study shifts the outcome of interest from sanctioning undemocratic behavior to preferences over \emph{delegation versus electoral accountability}.

Our core claim is that delegation is \emph{ideologically valenced}. When “experts” are perceived as aligned with incumbent parties, electoral losers may resist delegating authority away from elected politicians even if they support checks and balances. We derive two implications. First, the association between being farther from the governing bloc and support for delegation is conditional on the perceived ideological valence of expert institutions (negative when expert authority is government–aligned; potentially positive when neutral or loser–aligned). Second, this relationship should strengthen with individual partisan intensity and with higher system–level polarization \parencite{Svolik2020QJPS}.

Empirically, the analysis uses a novel survey of adults in Finland. Respondents reported closeness to each parliamentary party; these evaluations are combined into a respondent–level proximity to the governing coalition by weighting coalition parties by their parliamentary seat shares and renormalizing weights over the parties each respondent rated. The proximity measure is transformed into a normalized distance–to–government index. Support for technocratic delegation is measured on an ordered response scale. Models adjust for democratic satisfaction, trust in politicians, education, age, gender, and region.

The research design estimates linear specifications for ease of interpretation and ordered models as robustness. Results show a clear negative association between distance from the governing coalition and support for delegation. The pattern is robust to alternative distance constructions (seat–weighted, unweighted, closest–party) and to modeling choices. The evidence is consistent with expert governance being perceived as ideologically aligned with the incumbent bloc in this context: losers prefer electoral accountability over technocratic delegation. Exploratory heterogeneity indicates steeper negative slopes among respondents with stronger partisan attachments and in more polarized environments, with attenuation among those expressing higher trust in expert institutions. The findings refine scope conditions in the polarization literature by showing that losers’ consent does not automatically extend to technocracy and depends on the perceived ideological valence of delegated bodies.

\section{Theory}

Democratic governance typically balances technocratic delegation against democratic accountability. Technocratic delegation refers to entrusting policy authority to experts or non-partisan institutions insulated from direct electoral control. Examples include independent central banks, regulatory agencies, or constitutional courts staffed by specialists. Such delegation is justified by the promise of neutral competence and long-term policy stability, but it inherently trades off responsiveness to the electorate. In other words, empowering “the experts” may improve technocratic quality at the cost of reducing the public’s direct say in decision-making \parencite{Caramani2023}. This tension is well documented: technocratic policy-making can dilute traditional accountability, since decisions occur outside the arena of partisan competition and popular oversight \parencite{BertsouCaramani2020}. Democratic accountability, by contrast, requires that those who make policy are answerable to voters through elections and representative institutions. The crux of the issue is that what incumbents defend as neutral expertise, electoral losers may perceive as unaccountable elite rule.

A central claim of our theory is that electoral losers’ consent to delegation hinges on whether “neutral” institutions are seen as truly impartial or as ideologically aligned agents of the winners. Losers’ consent is crucial for democratic stability: those who lose an election must still accept the legitimacy of the system and trust that their interests will be protected in opposition \parencite{AndersonEtAl2005}. Typically, this is easier when independent institutions (courts, oversight bodies, expert agencies) serve as credible guardians of fair play. However, when losers suspect that these institutions are in fact partisan – stacked with or naturally sympathetic to the incumbent coalition’s ideology – their willingness to consent to delegation plummets. In effect, the losers no longer regard technocratic bodies as neutral arbiters but rather as extensions of the ruling bloc. Delegation under these conditions appears to them as a circumvention of democracy: policy being set by unelected actors who covertly advance the incumbents’ program. Losers will thus resist ceding authority to such bodies, invoking the need for accountability and the “will of the people” to counter what they view as a biased technocracy.

This argument is grounded in several streams of research. First, in polarized polities, even nominally non-partisan institutions become perceived in partisan terms. Polarization – the deep ideological divide between political camps – undermines any shared baseline of trust in institutions. \textcite{Svolik2019} demonstrates that intense polarization “undermines the public’s ability to serve as a democratic check”, as voters become willing to trade off democratic principles for partisan interests. In such an environment, each side suspects that ostensibly neutral referees (be it the judiciary, civil service, or expert regulators) covertly favor their opponents. Empirically, attitudes toward democratic institutions themselves have grown increasingly sorted by partisan identity. \textcite{ClaassenEtAl2024}, for example, show that in countries led by right-wing populists (Hungary, Poland, Brazil), supporters of the incumbent right exhibit markedly lower support for liberal-democratic checks and balances, whereas the opposing camp (the electoral “losers”) places greater value on those checks. The key insight is that polarization produces mirror-image perceptions: winners may dismiss certain constraints as nuisances, while losers champion them – but only so long as they believe those institutions can constrain the ruling side. If instead the institutions are seen as captured by the ruling side, this pattern flips. Under hyper-polarization, institutions that should function as neutral checks are readily turned into, or at least seen as, “unaccountable instruments of partisan or incumbent advantage”. In the words of \textcite{LiebermanEtAl2019}, polarization magnifies the partisan capture of independent institutions, eroding their legitimacy as neutral arbiters.

Given this dynamic, electoral losers will oppose technocratic delegation particularly when expert bodies carry an ideological valence aligned with incumbents. By ideological valence of expertise, we mean that expert institutions are not perceived as ideologically neutral but instead as tilting toward one end of the political spectrum. Expertise is never exercised in a political vacuum: economic technocrats, for instance, may predominantly subscribe to market-liberal orthodoxy; legal and bureaucratic elites might share the social outlook of the incumbent regime; scientific experts might prioritize values associated with one side of the cultural divide. In highly polarized contexts, these associations become highly salient. Losers come to view the “independent” central bank as effectively partisan (e.g. favoring austerity aligned with the government’s ideology), or the judiciary as packed with judges loyal to the ruling party. As a result, what might superficially be a technocratic policy – say, an impartial public-health directive or a monetary tightening – is interpreted through a partisan lens. Perceived ideological bias in technocratic institutions thus short-circuits losers’ trust. Rather than seeing delegation as securing unbiased expertise, losers see it as entrenching the incumbent coalition’s preferences behind a facade of objectivity.

Moreover, partisan actors often display a democratic hypocrisy in their support for institutions, endorsing or rejecting “neutral” oversight based on who benefits. Recent studies of U.S. public opinion underscore this point. Support for abstract democratic norms is high in principle but “softens considerably when they conflict with partisanship.” For instance, while 92% of Republicans supported robust congressional oversight of the executive under a Democratic president in 2022, only 65% did so under a co-partisan Republican president in 2019 – a swing driven by partisan calculus ￼. In general, both winners and losers tend to champion institutional checks when those serve their interests, and to attack or circumvent them when they do not \parencite{Brown2021}. This partisan double standard implies that an electoral loser’s professed desire for accountability may mask an underlying concern with policy outcomes. When their side is out of power, they decry “unelected bureaucrats” or “activist judges” only if those actors align with the rival camp. Our theory thus connects with work on partisan democratic backsliding: just as incumbents sometimes erode checks to entrench power, oppositions can withdraw loyalty from institutions they deem compromised. In both cases, commitment to democratic rules becomes contingent on who controls the levers of authority \parencite{WaldnerLust2018}.

We argue that opposition to technocratic delegation will be most acute under two conditions. First, when political polarization and partisan identity are strong. In a polarized environment, mutual mistrust runs high and even independent institutions are painted in red or blue. High affective partisan identification means citizens are more prone to believe conspiracy-esque narratives that “the experts” are in cahoots with their opponents. They are also less likely to tolerate policies made outside electoral contest, since compromise with the other side is seen as betrayal. Under such polarization, the basic requirement for losers’ consent – confidence that the system treats sides fairly – is in short supply. Second, when incumbents have in fact politicized or captured nominally technocratic institutions. While perceptions matter most for our mechanism, perceptions are often rooted in reality. Incumbent governments frequently use appointments and institutional reforms to bias ostensibly independent bodies in their favor, especially during episodes of democratic backsliding \parencite{LührmannLindberg2019}. Examples abound: packing supreme courts, tilting election commissions, staffing central bank boards with ideological fellow-travelers. These actions do not go unnoticed by the opposition. On the contrary, they become focal points of political conflict. An opposition that sees, say, the electoral commission openly favoring the ruling party will stridently object to any further empowerment of that commission. In new or fragile democracies, losing parties sometimes even refuse to recognize referees (for example, boycotting a “biased” election commission or constitutional court) precisely because they view them as tools of the incumbents. Thus, where the integrity of technocratic institutions is compromised, losers respond by rejecting those institutions’ authority.

Importantly, our theory does not claim that electoral losers inherently oppose expert rule per se. Rather, the support or opposition of losers toward technocratic delegation is conditional on the perceived neutrality of expertise relative to partisan competition. In contexts where expert institutions are broadly trusted across the spectrum, losers may willingly delegate decisions to them – indeed, historically, opposition parties often support strengthening independent authorities as a way to constrain powerful executives. Likewise, survey evidence suggests that citizens who feel alienated from partisan politics (so-called “structural losers”) can exhibit higher support for technocratic governance as an alternative to dysfunctional party competition \parencite{BertsouCaramani2020}. However, those scenarios presume that expertise stands apart from partisan ideology. Our contribution is to theorize what happens when that presumption breaks down. When expertise itself acquires an ideological hue, delegation no longer offers a neutral solution. To the contrary, it appears as a mechanism for the ruling coalition to lock in its preferences without democratic contest. Under such conditions, we predict that electoral losers will frame technocratic delegation as illegitimate and resist it vigorously. They will call instead for decisions to be returned to “the people” (often via elections or referendums) where the opposition has a fighting chance. In short, when expert institutions are ideologically aligned with incumbents, opposition actors interpret technocratic delegation as partisan entrenchment and respond by reasserting democratic accountability through popular control.

This theoretical logic yields several observable implications. We would expect, for instance, that losers’ resistance to delegation is heightened in highly polarized, ideologically sorted party systems, and during periods of intense policy conflict between government and opposition. We would also expect rhetorical parallels with populism: the opposition may adopt anti-elite, anti-expert language (“Let the people decide, not unelected experts!”) closely resembling populist discourse \parencite{KriesiPappas2015}. Indeed, our framework helps explain why populist movements (often composed of “losers” of globalization or of recent elections) frequently target technocratic institutions as part of a corrupt establishment \parencite{Caramani2023}. Conversely, in consensus-oriented or depolarized settings, or where technocratic bodies enjoy cross-partisan legitimacy, we should see less of this loser’s backlash. Ultimately, the theory underscores a paradox of contemporary democracy: institutions designed as impartial checks can sustain democratic stability only if they are broadly seen as impartial. If they come to be viewed as the ideological instruments of those in power, the very actors who most need to accept them – the electoral losers – may withdraw their consent, imperiling the delicate balance between expertise and democracy.


\section{Methods} 

\subsection*{1. Constructing \texttt{gov\_distance\_w\_01}}

The main independent variable is \texttt{gov\_distance\_w\_01}, a respondent–government distance measure based on party–closeness ratings.

Let $i$ index respondents and let $p \in G$ index parties in the governing coalition. From Q27, respondents rate their closeness to each parliamentary party on a $0$–$10$ scale (code $12$ = “don’t know”). Values equal to $12$ are set to missing.

Let $C_{ip} \in [0,10]$ denote respondent $i$’s closeness to party $p$. Let $s_p$ be party $p$’s seat count in the Eduskunta at the time of fieldwork (KOK $=48$, PS $=46$, RKP $=9$, KD $=5$), and define coalition seat weights
\[
w_p \equiv \frac{s_p}{\sum_{q \in G} s_q}\, .
\]
Because some respondents skip one or more government parties, weights are renormalized over the set of parties that respondent $i$ rated:
\[
\tilde w_{ip} \equiv 
\frac{w_p \,\mathbbm{1}\!\{C_{ip}\ \text{observed}\}}
{\sum_{q \in G} w_q \,\mathbbm{1}\!\{C_{iq}\ \text{observed}\}}\, .
\]
Seat–weighted coalition closeness is
\[
\text{GovCloseness}_i \equiv \sum_{p \in G} \tilde w_{ip}\, C_{ip} \in [0,10]\, .
\]
This is scaled to $[0,1]$ and complemented by a distance index:
\[
\texttt{gov\_closeness\_w\_01}_i \equiv \frac{\text{GovCloseness}_i}{10},
\qquad
\texttt{gov\_distance\_w\_01}_i \equiv 1 - \texttt{gov\_closeness\_w\_01}_i\, .
\]
By construction, larger \texttt{gov\_distance\_w\_01} indicates greater distance from the governing coalition in the Q27 closeness space.

\subsection*{2. Estimation model}

Support for technocratic delegation is modeled using a linear specification, treating the $1$–$5$ outcome as approximately continuous (ordered models are used as robustness):
\[
\begin{aligned}
\underbrace{\texttt{techno5}_i}_{\text{Technocracy}}
&= \beta_0
  + \beta_1\,\texttt{gov\_distance\_w\_01}_i
  + \beta_2\,\underbrace{\texttt{Q8\_1}_i}_{\text{Dem satisfaction}}
  + \beta_3\,\underbrace{\texttt{Q9\_4}_i}_{\text{Trust in Finnish politicians}} \\
&\quad + \boldsymbol{\gamma}'\,\mathbf{X}_i
  + \delta_{\text{region}(i)} + \varepsilon_i \, .
\end{aligned}
\]

The vector $\mathbf{X}_i$ contains education, age, and gender indicators; $\delta_{\text{region}(i)}$ are region fixed effects. The parameter of primary interest is $\beta_1$.

\FloatBarrier
%\bibliography{library}

\end{document}